---
title: "Chromosomal Instability Index Analysis"
author: "Lavinia Carabet"
output: 
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Chromosomal Instability Index (CIN) Analysis

CIN Analysis - Quantitative characterization of genome-wide copy number alterations as a measure of chromosomal instability at chromosome and cytoband levels between groups of patients

Case study: Colorectal Cancer (CRC)

Design: 38 Stage II CRC tissue samples, 15 with and 23 without relapse of tumor

Platform: Agilent Human Genome Comparative Genomic Hybridization (CGH) microarray
          aCGH - compares patients's genomes against a reference genome, identifies their differences and 
          locates regions of genomic imbalances in patients
          

Dataset: GEO GSE17181 <https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE17181>

## Clinical cohort selection

Prior to analysis clinical cohorts are selected based on the desired criteria, 
here having Event indicator for Disease Free Survival (EVENT_DFS) as outcome.

In other words, creating two groups:

- Relapse patients with Disease free survival as EVENT
- Relapse free patients with Disease free survival as CENSORING

```{r}
clinical.dat <- read.csv('./CRC_BROSEN_2010_01_CLINICAL_DATA.csv', header = T)

relapse.group <- clinical.dat[clinical.dat$EVENT_DFS == "EVENT",]
relapse.group

relapse.free.group <- clinical.dat[clinical.dat$EVENT_DFS == 'CENSORING',]
relapse.free.group

```

## Load chromosome-level CIN data matrix for analysis

```{r}
load("CRC_BROSENS_2010_01_CIN_CHROMOSOMES.Rda")

# dataMatrix - chromosome-level CIN matrix, each row is a chromosome, each column is a sample(patient)
dataMatrix[,1:3]

allids <- dimnames(dataMatrix)[[2]]

# relapse group IDs
grp1ids <-  allids[1:15]
# relapse free group IDs
grp2ids <- allids[16:38]

sampleGrp1Len <- length(grp1ids)
sampleGrp2Len <- length(grp2ids)

## clinical.inf: n*2 matrix, the 1st column is 'sample name', the second is 'label'
clinical.inf <- as.matrix( cbind( dimnames(dataMatrix)[[2]], 
                                  c( rep('Relapse Free', sampleGrp2Len), 
                                     rep('Relapse',sampleGrp1Len ) ) ) )
clinical.inf
```

## CIN Chromosome level analysis

```{r, fig.height=8, fig.width=8}
source("getSubmatrix.R")
cinInputMatrix <- getSubmatrix.twogrps(dataMatrix, grp2ids, grp1ids)

source("heatmap.draw.R")
heatmap.draw(cinInputMatrix, clinical.inf)
```

## CIN Cytoband level analysis

```{r}
load("CRC_BROSENS_2010_01_CIN_CYTOBANDS.Rda")

# dataMatrix - cytoband-level CIN matrix, each row is a cytoband, each column is a sample(patient)
head(dataMatrix[,1:3],20)
tail(dataMatrix[,1:3],20)

allids <- dimnames(dataMatrix)[[2]]

grp1ids <-  allids[1:15]
grp2ids <- allids[16:38]

sampleGrp1Len <- length(grp1ids)
sampleGrp2Len <- length(grp2ids)

## clinical.inf: n*2 matrix, the 1st column is 'sample name', the second is 'label'
clinical.inf <- as.matrix( cbind( dimnames(cinInputMatrix)[[2]], 
                                  c( rep('Relapse Free', sampleGrp2Len), 
                                     rep('Relapse',sampleGrp1Len ) ) ) )
clinical.inf
```

```{r, fig.height=8, fig.width=8}
cinInputMatrix <- getSubmatrix.twogrps(dataMatrix, grp2ids, grp1ids)

load("hg18_annot.Rda")                  # load reference genome, human genome 18 data frame
annotInputMatrix <- hg18_annot          # each row corresponds to a cytoband
head(annotInputMatrix,5)                # columns: "chrom", "start", "end", "name", and "stain"

source("cytobands_cin.draw.R")
#for (i in 1:22) {
for (chr in c(4,18)) {
  cytobands_cin.draw(cinInputMatrix, clinical.inf, chr, annotInputMatrix, 
                     title_text = paste('chromosome',chr,'cytobands CIN overview'))
}
```

Above are the results of CIN analysis showing cumulative instabilities (losses and gains) at chromosome and 
cytoband levels with more instabilities in the relapsed group compared to relapse-free in chromosomes 4 and 18 

## Downstream T-test Cytoband-level Differential Analysis

CRC_BROSENS_2010_01_CIN_CYTOBANDS.Rda cytoband CIN data and groups above

```{r}
# function to calculate Student’s two-sample t-test on all cytobands
# function returns the p-value for the test
# NAs are removed for each test

t.test.all.cyt <- function(x,s1,s2) {
	x1 <- x[s1]
	x2 <- x[s2]
	x1 <- as.numeric(x1)
	x2 <- as.numeric(x2)
	t.out <- t.test(x1, x2, alternative='two.sided', var.equal = TRUE)
	out <- as.numeric(t.out$p.value)
	return(out)
}

t.test.run <- apply(cinInputMatrix, 1, t.test.all.cyt, s1=grp1ids, s2=grp2ids)

# calculate means of the groups 
grp1.mean <- apply(cinInputMatrix[,grp1ids], 1, mean, na.rm = T)
grp2.mean <- apply(cinInputMatrix[,grp2ids], 1, mean, na.rm = T)

# calculate fold change
fold.change <- grp1.mean - grp2.mean 
#range(fold.change)

#Build data frame with p-value and fold information
ds.featured.cyt <- cbind(t.test.run, fold.change)

# Select cytobands with significance alpha=0.05 and fold.change = 1
featured.sel.cyt <- (t.test.run < 0.05) & ((fold.change < -log2(1)) | (fold.change > log2(1)))

#Filters data frame with featured selection cytobands found
diff.changed.cyt <- ds.featured.cyt[featured.sel.cyt,]

#Orders data frame by fold change
diff.changed.cyt <- diff.changed.cyt[order(diff.changed.cyt[,2], decreasing = T),]
diff.changed.cyt
```

The follow-up T-test differential analysis at cytoband-level showed that the most differentially changed cytobands are in chromosomes 4 and 18 (as well as 3), correlating with results from the literature.

Chromosomal instability in 4q arm (4q22.2-35.2) has been previously reported to be associated with local recurrence in colon cancer and worst outcomes in stage II CRC patients. 

In addition, the cytoband region 18q21.2 contains Deleted in Colorectal Carcinoma (DCC) gene where frequent loss of heterozygosity (LOH) events in colon cancer occur, as per OMIM (Online Mendelian Inheritance in man). 



