---
title: "scRNA-seq"
author: "Lavinia Carabet"
date: "08/26/2021"
output: 
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Loading

Read the 10X Genomics PMBC scRNA-seq expression sample data stored in the raw_feature_bc_matrix directory, using the Seurat package.

For downloading the data use `scRNA-seq.ipynb`

The data encompasses:

* one barcode file (barcodes.tsv.gz),
* one feature file (features.tsv.gz), and
* one expression matrix file (matrix.mtx.gz)

```{r, message=FALSE, warning=FALSE}
#install.packages("Seurat")
```
```{r, message=FALSE, warning=FALSE}
library(Seurat)

data.dir <- 'raw_feature_bc_matrix/'
list.files(data.dir)

pbmc <- Read10X(data.dir = data.dir)
pbmc[1:10, 1:5]
 
pbmc <- CreateSeuratObject(counts = pbmc, min.cells = 3, min.features = 30)
pbmc
```
```{r}
# number of features and cells in the active assay
dim(pbmc)
head(rownames(pbmc)) # features
head(colnames(pbmc)) # cells
# specific assay
names(pbmc)
# pulling specific Assay
pbmc[['RNA']]

GetAssayData(pbmc)[1:10, 7:10]

# cell-level metadata data frame
colnames(pbmc[[]])
head(x = pbmc[[c('nCount_RNA', 'nFeature_RNA')]])
```
## Data Preprocessing Workflow

### 1. Quality Control (QC) and Cell Filtering

QC metrics

* check for empty droplets that can lead to few detected genes, `nFeature_RNA`.
`nFeature_RNA` is the number of genes detected in each cell. `Low nFeature_RNA` for a cell indicates that it may be dead/dying or an empty droplet

* check for cell doublets or multiplets that can lead to abnormally high gene count, `nCount_RNA`.
`nCount_RNA` is the total number of molecules detected within a cell. `High nCount_RNA` indicates that a cell may be a doublet (or multiplet)

* check for extensive mitochondrial contamination due to cell debris and dying cells, computed percentage of transcripts that map to mitochondrial genes `percent_mt`. Filter out low-quality cells having `high percent_mt`, high percent of counts originating from the set of mitochondial genes (features) starting with `MT-`

```{r}
# compute the percentage of transcripts that map to mitochondrial genes
pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")
head(pbmc[[]])
```

Visualize the distribution of the parameters
```{r}
# visualize QC metrics as a violin plot
VlnPlot(pbmc, features = c('nFeature_RNA','nCount_RNA', 'percent.mt'), ncol = 3)
```
```{r, fig.width=12, fig.height=4}
# visualize feature-feature relationships using FeatureScatter
# Pearson correlation between two features displayed above the plot
plot1 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot2 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot1 + plot2
```
Filter cells
- with less than 200 unique gene/feature counts
- with greater than 25000 unique molecular identified (UMI) feature counts per cell
- with greater than 15% mitochondrial counts
```{r}
pbmc <- subset(pbmc, subset = nFeature_RNA > 200 & nCount_RNA < 25000 & percent.mt < 15)
```
### 2. Data Normalization

`LogNormalize` method normalizes the feature expression measurements for each cell by the total expression, multiplies this by a scale factor (10,000 by default), and log-transforms the result
```{r}
pbmc <- NormalizeData(pbmc, normalization.method = "LogNormalize", scale.factor = 10000)
```
```{r}
# normalized data
pbmc[["RNA"]]@data[1:10, 1:5]
```
### 3. Feature Selection

`FindVariableFeatures` function to compute a subset of features that exhibit high cell-to-cell variation in the dataset (high/low expression in various cells) for downstream analysis
```{r}
pbmc <- FindVariableFeatures(pbmc, selection.method = "vst", nfeatures = 2000)

# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(pbmc), 10)
```
```{r, message=FALSE, warning=FALSE}
# plot variable features 
plot1 <- VariableFeaturePlot(pbmc)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE, xnudge = 0, ynudge = 0)
plot2
```

### 4. Data Scaling

`ScaleData` function to linearly transform the data by shifting and scaling the expression of each gene to mean expression 0 and variance 1 across all cells, giving equal weights to genes so that highly-expressed genes do not dominate in downstream analyses
```{r}
pbmc <- ScaleData(pbmc, features = rownames(pbmc))
# scaled data
pbmc[["RNA"]]@scale.data[1:10, 1:5]
```
## Dimensionality Reduction

### Linear Principal Component Analysis

`RunPCA` using only the 2000 selected and scaled highly variable features (out of 15262 total features)
```{r}
pbmc <- RunPCA(pbmc, features = VariableFeatures(pbmc))
```
Examine the PCA results beyond defaults (ndims.print = 10 instead of 5 out of total npcs =50 computed, nfeatures.print = 10 instead of 30)
```{r}
print(pbmc[["pca"]], dims = 1:10, nfeatures = 10)
```

Visualize PCA results
```{r, fig.height=6, fig.width=10}
VizDimLoadings(pbmc, dims = 1:2, reduction = "pca")
```
```{r, fig.height=10, fig.width=10}
DimHeatmap(pbmc, dims = 1:9, cells = 200, balanced = TRUE)
```

### Dimensionality Selection 

`ElbowPlot` function ranking of principle components based on the percentage of variance explained by each one
```{r, fig.height=4, fig.width=6}
ElbowPlot(pbmc)
```
`JackStraw` procedure for identifying significant PCs having a strong enrichment of features with low p-values 
```{r}
pbmc <- JackStraw(pbmc, num.replicate = 100)
pbmc <- ScoreJackStraw(pbmc, dims = 1:20)
```

```{r, message=FALSE, warning=FALSE, fig}
# comparing the distribution of p-values for each PC (solid curves) 
# with a uniform distribution (dashed line)
JackStrawPlot(pbmc, dims = 1:20, cols = DiscretePalette(20, "polychrome"), ymax = 1, xmax=1)
```

## Cluster Analysis

Cluster cells based on their expression pattern by first constructing a KNN graph based on the euclidean distance in PCA space using the `FindNeighbors` function with the first 10 most significant principal components PCs as input, and then iteratively group cells together using `FindClusters` function that implements Louvain modularity optimization.
```{r}
pbmc <- FindNeighbors(pbmc, dims = 1:10)
pbmc <- FindClusters(pbmc, resolution = 0.5)
```

Cluster IDs for the first 5 cells
```{r}
head(Idents(pbmc), 5)
```

## t-SNE Non-linear Dimensionality Reduction

t-Distributed Stochastic Neighbor Embedding (t-SNE) aims to learn the underlying manifold of the data in order to place similar cells together in low-dimensional space.  Cells within the graph-based clusters should co-localize on dimension reduction plots.
`RunTSNE` function using same PCs as input as for cell clustering
```{r}
pbmc <- RunTSNE(pbmc, dims = 1:10, tsne.method = "Rtsne")
```

### Visualize scRNA-seq cell clusters using t-SNE

using `TSNEPlot` function or `DimPlot` function
```{r}
TSNEPlot(pbmc, label = TRUE)
#DimPlot(pbmc, reduction = "tsne", label = TRUE)
```
### Visualize number of detected genes (nFeature_RNA) and number of UMIs (nCount_RNA) on t-SNE 
```{r, fig.height=3, fig.width=8}
 FeaturePlot(pbmc, features = c("nFeature_RNA", "nCount_RNA"), 
             cols = c("grey", "lightblue"), pt.size = 0.5, reduction = "tsne")
```

## Save the Seurat object
```{r}
names(pbmc)
saveRDS(pbmc, file = "./output/pbmc_out.rds")
```
## Identify Cluster Biomarkers via Differentially Expressed Features 

`FindAllMarkers` function finds biomarkers for each cluster compared to all remaining cells, reporting both positive and negative markers by default.

It supports a number of differential expression (DE) tests through `test.use` argument:

* `wilcox` : Wilcoxon rank sum test (default)

* `bimod` : Likelihood-ratio test for single cell feature expression

* `roc` : Standard AUC classifier

* `t` : Student’s t-test

* `poisson` : Likelihood ratio test assuming an underlying negative binomial distribution. Use only for UMI-based datasets

* `negbinom` : Likelihood ratio test assuming an underlying negative binomial distribution. Use only for UMI-based datasets

* `LR` : Uses a logistic regression framework to determine differentially expressed genes. Constructs a logistic regression model predicting group membership based on each feature individually and compares this to a null model with a likelihood ratio test.

* `MAST` : Model-based Analysis of Single-cell Transcriptomics - Generalized Linear Model (GLM)-framework that treats cellular detection rate as a covariate, specially adapted for bimodal and/or zero-inflated single cell gene expression data

* `DESeq2` : DE based on a model using the negative binomial distribution

### Wilcox test
```{r, message=FALSE, warning=FALSE}
library(dplyr)
```
```{r}
pbmc.markers.wilcox <- FindAllMarkers(pbmc, only.pos=FALSE, test.use = "wilcox", 
                                      min.pct = 0.25, logfc.threshold = 0.25)
head(pbmc.markers.wilcox)
top10.per.cluster <- pbmc.markers.wilcox %>% group_by(cluster) %>% top_n(n = 10, wt = avg_log2FC)
top10.per.cluster
```
### ROC test

Returns the `classification power` for any individual biomarker (ranging from 0 - random, to 1 - perfect)
```{r}
pbmc.markers.roc <- FindAllMarkers(pbmc, logfc.threshold = 0.25, min.pct = 0.25,
                                   test.use = "roc", only.pos = FALSE)
head(pbmc.markers.roc)
top10.per.cluster.roc <- pbmc.markers.roc %>% group_by(cluster) %>% top_n(n = 10, wt = avg_log2FC)
top10.per.cluster.roc
```
### DESeq2 test

```{r, message=FALSE, warning=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("DESeq2")
```

```{r}
pbmc.markers.deseq2 <- FindAllMarkers(pbmc, logfc.threshold = 0.25, min.pct = 0.25, 
                                      test.use = "DESeq2", only.pos = FALSE, max.cells.per.ident = 100)
head(select(pbmc.markers.deseq2, -cluster))
```

### Visualize top10 Wilcox DE biomarkers for each cluster as an expression heatmap

```{r, fig.height=10, fig.width=10}
DoHeatmap(pbmc, features = top10.per.cluster$gene) 
```

## Visualize top10 ROC biomarkers for each cluster as an expression heatmap

```{r, fig.height=10, fig.width=10}
DoHeatmap(pbmc, features = top10.per.cluster.roc$gene) 
```

### Visualize expression level of cluster specific genes/features on t-SNE using FeaturePlot
```{r, fig.height=6, fig.width=8}
# FeaturePlot(pbmc, 
#             features = c("CD14", "LYZ", "IL7R", 
#                          "CCR7", "CST3", "CD8A",
#                          "MS4A1", "CD79A", "NKG7", 
#                          "GNLY", "FCGR3A", "PPBP"),
#             cols = c("grey", "orange"), pt.size = 0.5, reduction = "tsne")
FeaturePlot(pbmc, 
            features = c("CD14", "LYZ", "IL7R", "CCR7"),
            cols = c("grey", "orange"), pt.size = 0.5, reduction = "tsne")

FeaturePlot(pbmc, 
            features = c("CST3", "CD8A", "MS4A1", "CD79A"),
            cols = c("grey", "orange"), pt.size = 0.5, reduction = "tsne")

FeaturePlot(pbmc, 
            features = c("NKG7", "GNLY", "FCGR3A", "PPBP"),
            cols = c("grey", "orange"), pt.size = 0.5, reduction = "tsne")
```

### Visualize expression level of top10 most variable genes on t-SNE 
```{r, fig.height=6, fig.width=8}
# FeaturePlot(pbmc, 
#             features = top10,
#             cols = c("grey", "salmon"), pt.size = 0.5, reduction = "tsne")

FeaturePlot(pbmc, 
            features = top10[1:4],
            cols = c("grey", "salmon"), pt.size = 0.5, reduction = "tsne")

FeaturePlot(pbmc, 
            features = top10[5:8],
            cols = c("grey", "salmon"), pt.size = 0.5, reduction = "tsne")
```
```{r, fig.height=4, fig.width=8}
FeaturePlot(pbmc, 
            features = top10[9:10],
            cols = c("grey", "salmon"), pt.size = 0.5, reduction = "tsne")
```

## Assign Cell Type Identity to Clusters
```{r}
clusters <- levels(pbmc)
markers <- c("CD14", "IL7R", "CCR7", "CST3", "CD8A", "MS4A1", "CD79A", "NKG7")
cell.types <- c("CD14+ Mono", "Memory CD4+ T", "Naive CD4+ T", "DC", "CD8+ T", "B", "B-like", "NK" )
cluster.celltype.mapping <- as_tibble(cbind(clusters, markers, cell.types))
cluster.celltype.mapping
names(cell.types) <- clusters


pbmc.final <- RenameIdents(pbmc, cell.types)
DimPlot(pbmc.final, reduction = "tsne", label = TRUE, pt.size = 1) + NoLegend()
```

```{r}
saveRDS(pbmc, file = "./output/pbmc_final_out.rds")
```


```{r}
sessionInfo()
```