---
title: "Copy number estimation and CNA detection in tumor samples from WGS HTS data using HMMCopy"
author: "Lavinia Carabet"
date: "11/19/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Setup the libraries and input data
### Install libraries
```{r, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("HMMcopy")
```

### Load libraries
```{r, message=FALSE, warning=FALSE, results="hide"}
options(stringsAsFactors = TRUE)
library(HMMcopy)
```
## 2. Generating Copy Number Profiles

### Load normal HTS copy number data
```{r}
rfile <- system.file("extdata", "normal.wig", package = "HMMcopy")
gfile <- system.file("extdata", "gc.wig", package = "HMMcopy")
mfile <- system.file("extdata", "map.wig", package = "HMMcopy")
normal_reads <-  wigsToRangedData(rfile, gfile, mfile)
normal_reads[1000:1010,]
#dim(normal_reads)
```

### Correcting normal HTS copy number data
```{r, warning=FALSE}
normal_copy <-  correctReadcount(normal_reads)
normal_copy[1000:1010,]
```

### Visualizing the effects of correction
```{r}
par(cex.main = 0.7, cex.lab =0.7, cex.axis =0.7, mar = c(4,4,2, 0.5))
plotBias(normal_copy, pch =20, cex = 0.5)
```

### Visualizing corrected copy number profiles
```{r, warning=FALSE}
par(mar = c(4,4,2, 0))
plotCorrection(normal_copy, pch = ".")
```

### Correcting and visualizing tumor copy number profiles
```{r, warning=FALSE}
tfile <- system.file("extdata","tumour.wig", package = "HMMcopy")
tumor_copy <- correctReadcount(wigsToRangedData(tfile, gfile, mfile))
par(mar = c(4,4,2, 0))
plotCorrection(tumor_copy, pch = ".")
```

## 3. Segmentation and Classification of Copy Number Profiles
```{r, warning=FALSE}
tumor_segments <- HMMsegment(tumor_copy)
```
### Visualizing segments and classified states
```{r}
par(mfrow = c(2,1), cex.main = 0.5, cex.lab = 0.7, cex.axis =0.7, 
    mar = c(4, 4,0,0),  mgp = c(1,0.5,0))
plotSegments(tumor_copy, tumor_segments, pch=".", 
             ylab = "Tumor Copy Number", xlab = "Chromosome Position")
cols <- stateCols() #6 default state colors
legend("topleft", c("HOMD", "HETD", "NEUT", "GAIN", "AMPL", "HLAMP"), 
       fill = cols, horiz = TRUE, bty = "n", cex = 0.7)
```

### Improving segmentation performance
```{r, warning=FALSE}
default_param <- HMMsegment(tumor_copy, getparam = TRUE)
default_param
#6 states, 10 parameters matrix
```

### Reducing the number of segments a.k.a increasing the length of segments
```{r, warning=FALSE}
longseg_param <-  default_param
longseg_param$e <-  0.9999999999999999
longseg_param$strength <- 1e30
longseg_segments <-  HMMsegment(tumor_copy, longseg_param)
```
### Visualizing segments and classified states to confirm decrease in segments as intended
```{r}
par(mfrow = c(2,1), cex.main = 0.5, cex.lab = 0.7, cex.axis =0.7, 
    mar = c(4, 4,0,0),  mgp = c(1,0.5,0))
plotSegments(tumor_copy, longseg_segments, pch=".", 
             ylab = "Tumor Copy Number", xlab = "Chromosome Position")
cols <- stateCols() #6 default state colors
legend("topleft", c("HOMD", "HETD", "NEUT", "GAIN", "AMPL", "HLAMP"), 
       fill = cols, horiz = TRUE, bty = "n", cex = 0.7)
```

### Adjusting copy number state ranges - correcting the incorrect median of each copy number state in the plot above
### problem with mu parameter
```{r}
#output of segmentation process: matrix of the median of 6 states (rows) after each iteration (7) 
#of the optimization algorithm (columns)
longseg_segments$mus
```
### first column initial suggested mu
```{r}
longseg_param$mu
```

### last column actual values used during the segmentation process
```{r}
longseg_segments$mus[,7]
```

### Visualising the problem - Medians not running through middle of segments of many states

```{r}
par(mfrow = c(2,1), cex.main = 0.5, cex.lab = 0.7, cex.axis =0.7, 
    mar = c(4, 4,0,0),  mgp = c(1,0.5,0))
plotSegments(tumor_copy, longseg_segments, pch=".", 
             ylab = "Tumor Copy Number", xlab = "Chromosome Position")
for (i in 1:nrow(longseg_segments$mus)) {
  abline(h=longseg_segments$mus[i, ncol(longseg_segments$mus)], col = cols[i], lwd =2, lty =3)
}
abline(v = 7.68e7, lwd = 2, lty = 3)
abline(v = 8.02e7, lwd = 2, lty = 3)
```

### Update/Re-initialiaze mu param to solve the problem 

```{r, warning=FALSE}
newmu_param <- longseg_param
newmu_param$mu <-  c(-0.5, -0.4,-0.15, 0.1,0.4, 0.7) # why these values? nothing has changed
newmu_segments <- HMMsegment(tumor_copy, newmu_param)
par(mfrow = c(2,1), cex.main = 0.5, cex.lab = 0.7, cex.axis =0.7, 
    mar = c(4, 4,0,0),  mgp = c(1,0.5,0))
plotSegments(tumor_copy, newmu_segments, pch=".", 
             ylab = "Tumor Copy Number", xlab = "Chromosome Position")
# for (i in 1:nrow(newmu_segments$mus)) {
#   abline(h=newmu_segments$mus[i, ncol(newmu_segments$mus)], col = cols[i], lwd =2, lty =3)
# }
#  abline(v = 7.68e7, lwd = 2, lty = 3)
#  abline(v = 8.02e7, lwd = 2, lty = 3)
```


### Understanding parameter convergence 
```{r}
newmu_segments$mus
#newmu_param$mu <-  c(-0.5, -0.4,-0.15, 0.1,0.4, 0.7) #alg ignores these param values
```
### Understanding parameter convergence 
```{r}
longseg_param$mu
```
### Overriding parameter convergence 
```{r, warning=FALSE}
#solution disallow the alg from making large shifts to mu
#achieved by setting the prior mean of n (i.e.m - optimal value of mu) to values identical to mu
par(mfrow = c(2,1), cex.main = 0.5, cex.lab = 0.7, cex.axis =0.7, 
    mar = c(4, 4,0,0),  mgp = c(1,0.5,0))
newmu_param$m <- newmu_param$mu
realmu_segments <- HMMsegment(tumor_copy, newmu_param)
plotSegments(tumor_copy, realmu_segments, pch=".", 
             ylab = "Tumor Copy Number", xlab = "Chromosome Position")
# for (i in 1:nrow(realmu_segments$mus)) {
#   abline(h=realmu_segments$mus[i, ncol(realmu_segments$mus)], col = cols[i], lwd =2, lty =3)
# }
#  abline(v = 7.68e7, lwd = 2, lty = 3)
#  abline(v = 8.02e7, lwd = 2, lty = 3)
cols <- stateCols() #6 default state colors
legend("topleft", c("HOMD", "HETD", "NEUT", "GAIN", "AMPL", "HLAMP"), 
       fill = cols, horiz = TRUE, bty = "n", cex = 0.7)
```

## 4. Matched Tumor-Normal Sample Correction
### Normalizing tumor by normal copy number profiles
```{r}
somatic_copy <- tumor_copy
#LOGARITHM IDENTITY: log(a) -log(b) == log(a/b)
somatic_copy$copy <- tumor_copy$copy - normal_copy$copy
```
### Segmentation and visualization of somatic copy number aberration
```{r, warning=FALSE}
somatic_segments <- HMMsegment(somatic_copy, newmu_param) 
par(mfrow = c(2,1), cex.main = 0.5, cex.lab = 0.7, cex.axis =0.7, 
    mar = c(4, 4,0,0),  mgp = c(1,0.5,0))
plotSegments(somatic_copy, somatic_segments, pch=".", 
             ylab = "Tumor Copy Number", xlab = "Chromosome Position")
```

\pagebreak
### Export somatic copy number aberration
```{r}
#somatic_copy
somatic_segments$segs
#str(somatic_segments)
readr::write_delim(somatic_segments$segs,"somatic_segments_LAC.txt")
```
### Session info
```{r}
toLatex(sessionInfo())
```