---
title: "sparklyr - R interface to Apache Spark"
output: html_notebook
author: LAC
---

## Install

### Install sparklyr package

```{r}
install.packages("sparklyr")
```

### Install Spark locally

Apache Spark is a multi-language engine for executing data engineering, data science, and machine learning on single-node machines or clusters.

```{r}
library(sparklyr)

spark_install(version = "3.1.1")
```

```{r}
spark_installed_versions()
```
### Connect to the single node local Spark cluster 

`spark_connect()` returns a Spark Context which represents the connection to the Spark cluster used to execute Spark commands

```{r}
library(sparklyr)
sc <- spark_connect(master="local")
```

## Using Spark

```{r}
library(dplyr)
head(iris)
```

### Create a Spark DataFrame by copying iris dataset into Spark

```{r}
iris_sdf <- sdf_copy_to(sc=sc, x=iris, overwrite = T)
```

### Check data frames available in Spark

```{r}
src_tbls(sc)
```

```{r}
iris_sdf %>% count
```

```{r}
head(iris_sdf)
```
### Monitoring and analyzing execution of Spark commands through Spark's web interface

```{r, warning=FALSE, message=FALSE}
spark_web(sc)
```

### Fit a linear regression model to SparkDataFrame in Spark environment

Within Spark system there are many statistical and machine learning algorithms developed to run parallel across many CPUs with data distributed across many memory units. Through `sparklyr` package, many Spark Machine Learning Library (MLlib) algorithms can be accessed in R. These include: linear regression, survival regression. generalized linear regression, naive-bayes, decision trees, random forests, gradient-boosted trees, multilayer perceptron, principal component analysis, k-means clustering, etc.

```{r, warning=FALSE, message=FALSE}
partitions <- iris_sdf %>%
  sdf_random_split(training = 0.8, test = 0.2, seed = 2222)

lm_model <- partitions$training %>%
  ml_linear_regression(Sepal_Length ~ Sepal_Width + Petal_Length + Petal_Width)

summary(lm_model)
```

```{r}
lm_pred <- ml_predict(lm_model, partitions$test)
head(lm_pred,6)
```

```{r}
# calculate RMSE as performance metric for the regression model
# the closer RMSE is to 0, the more accurate the model's predictive power
ml_regression_evaluator(lm_pred, label_col = "Sepal_Length") 
```

### Fit a Multilayer Perceptron Classification model 

Classification model based on the Artificial Neural Networks - Multilayer Perceptron. Each layer has sigmoid activation function, output layer has softmax.
```{r}
mlp_model <- partitions$training %>%
  ml_multilayer_perceptron_classifier(Species ~ ., layers = c(4,3,3))

mlp_pred <- ml_predict(mlp_model, partitions$test)

head(mlp_pred,10)
```

```{r}
# calculate F1-score as performance metric for the multiclass classification model
# F1 ranges from 0 to 1. The closer to 1 (high F1 - high precision and recall), the better the model
ml_multiclass_classification_evaluator(mlp_pred)
```